---
title: "Summary Advice Sheet for blackspot seabream in GSAs 1 and 3 (SBR 1,3)"

author: 
 - WGSAD-GFCM 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    number_sections: yes
    #toc: yes
    #keep_tex: false
vignette: >
  %\VignetteIndexEntry{ss3daigs Vignetted}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r wrap-hook, echo = FALSE,message=FALSE,warning=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
```


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "  " ,fig.align = 'center', cache=FALSE,tidy.opts=list(width.cutoff=80), tidy=TRUE)

```

# Background

This vignette present the analysis to produce the output summary of the stock status for blackspot seabream in the GSAs 1 and 3 (GSA-1-3), including short- and long-term forecasts.


## Load packages


```{r,message=F,warning=F}
library(r4ss)
library(ss3diags)
library(FLCore)
library(FLBRP)
library(FLasher)
library(FLSRTMB)
library(ggplotFL)
library(FLRef)
library(ggplot2)
library(ss3om)
library(FLRef)
```

# Stock Status

Load reference case model


```{r}
model = "basecase"
```

Load reference model 

```{r,eval=F}

ss3rep = SS_output(model)

dir.create("rdata",showWarnings = F)
save(ss3rep,file="rdata/SBR_1_3_RefYr2023.rdata")

```

```{r}
# 2023  reporting (data) year - reference case 
load(file="rdata/SBR_1_3_RefYr2023.rdata",verbose=T)
ref2023 = ss3rep
years=1983:2023

mvnbase= FLRef::ssmvln(ref2023,addprj = F, Fref="Btgt",verbose=T)
```


```{r}
sstk = ss2FLStockR(mvnbase)
# Add Bpa + Blim
sstk@refpts = rbind(sstk@refpts,
                FLPar(Bpa=sstk@refpts[[2]]*0.5,Blim=sstk@refpts[[2]]*0.25)) 
# with uncertainty
stki = ss2FLStockR(mvnbase,output="iters",thin=1) 
stki@refpts = sstk@refpts= sstk@refpts[c(1,2,6,7)]
```

```{r fig31, warning=F,message=F, fig.height=7,fig.width=9, fig.cap = paste0("Estimated stock status trajectories with associated reference points for the 2024 base-case scenario of blackspot seabream in GSAs 1-3")}

plotAdvice(FLStocks(CIs=stki,Fit=sstk))+
   scale_fill_manual(values = c("dodgerblue","black"))+
  scale_color_manual(values = c(0,"black"))+
  theme(legend.position = "none")+scale_x_continuous(breaks=seq(min(years),max(years),2))+
  theme(axis.text.x = element_text(size=8, angle=90))

```



# F-based forecasting 


Load all runs in one go with `SSgetoutput`

```{r}
Ffrac = c(0.001,0.25,0.5,0.75,1,1.25)
runs = c(paste0("Fsq",Ffrac),"Ftgt")
fdirs = file.path("forecast",runs)

```

```{r,eval=F}
fcs = SSgetoutput(dirvec = fdirs)
save(fcs,file="rdata/forecasts.rdata")
```

```{r,echo=F}
load("rdata/forecasts.rdata")
```

Create list with outputs from `SSdeltaMVLN`

\pagebreak

Convert to `FLR` format


```{r}

fstks = FLStocks(Map(function(x,y){
  out = FLRef::ssmvln(x,Fref="Btgt",verbose=F,run=y,addprj=T)
  out = window(ss2FLStockR(out,thin=1),start=2023)
  
  out@refpts =  sstk@refpts # Remove B0
  return(out)
},x=fcs,y=as.list(c(paste0(Ffrac,"Fcur"),"Ftgt"))))
names(fstks) = c(paste0(c(0,Ffrac[-1]),"Fcur"),"Ftgt")

```

\pagebreak

```{r p10, fig.height=6.5,fig.width=9, warning=F,message=F, fig.cap = paste0("Trajectories for the basecase run and forecast scanarios relative $F_{cur} = F_{2023}$ and for $F_{tgt}$ ")}

# add ref
pstks = FLStocks(c(FLStocks(Assessment = sstk), fstks))
plotAdvice(window(pstks)) + geom_vline(xintercept = c(2023, 2024, 2030),
linetype = 2) + scale_color_manual(values = c("black", sscol(length(fstks)))) +
scale_x_continuous(breaks = seq(1985, 2045, 5)) + theme(axis.text.x = element_text(size = 8,
angle = 90))

```


\pagebreak

```{r}
out.st  =FLRef::fwd2stars(fstks,eval.yrs = 2026,rel=T)
out.lt = FLRef::fwd2stars(fstks,eval.yrs = 2042,rel=T)


knitr::kable(out.st,"pipe",
      align ="lccccc",caption="Summary of short-term forecast scenario results for 2026")

knitr::kable(out.lt,"pipe",
      align ="lccccc",caption="Summary of long-term forecast scenario results for 2042")

```


Save STAR data

```{r}
 
out = FLRef::flr2stars(stki)$timeseries
mles = FLRef::flr2stars(sstk)
# replace medians with mles (TODO: automise)
out$Rec=mles$timeseries$Rec
out$SSB=mles$timeseries$SSB
out$Bratio=mles$timeseries$Bratio
out$F=mles$timeseries$F
out$Fratio=mles$timeseries$Fratio


write.csv(out,file="sbr1.3.ts.stars.csv",row.names = F)
write.csv(mles$refpts,file="sbr1.2.3.refpts.stars.csv",row.names = F)

# Forecast

fwdout= cbind(out.st,out.lt[-1])
write.csv(fwdout,file="sbr1.3.fwd.stars.csv",row.names = F)


```


# ss3om

## Build FLStock

SS3 outputs are loaded with the `readFLSss3()` into an `FLStock` object. The folder that contains the model outputs has to be specified.

In the following, the area outside is evaluated first.

```{r}
run = "basecase"
stk = window(ss3om::readFLSss3(run,wtatage = TRUE,forecast=T),end=2023)
sr = ss3om::readFLSRss3(run)
stk@name = "SBR.1_3"
stk@desc = "2024, GFCM, SS3"
```

Check that the fbar range is consistent with `ss.starter` input.

```{r}
range(stk)

```
```{r fig8a, message=FALSE,warning=FALSE,fig.height=10,fig.width=6, fig.cap = paste0("Seasonal  stock trajectories for SBR 1-3") }
plot(stk)+theme_bw()+ylab("F")+xlab("Year")

```

\pagebreak


```{r fig8b, message=FALSE,warning=FALSE,fig.height=7,fig.width=8, fig.cap = paste0("Combined Stock trajectories for SBR 1-3") }
flqs= FLQuants(SSB=unitSums(ssb(stk)),Vuln.bio=unitSums(vb(stk)),Tot.bio=unitSums(stock(stk)),
               H=unitSums(catch(stk))/unitSums(vb(stk)))
plot(flqs)+theme_bw()+xlab("Year")+facet_wrap(~qname,scales="free_y")

```


```{r fig8c , message=FALSE,warning=FALSE,fig.height=7,fig.width=8, fig.cap = paste0("Combined Stock trajectories by sex for SBR 1-3") }
flqs= FLQuants(SSB=(ssb(stk[,,1])),Vuln.bio=(vb(stk)),Tot.bio=(stock(stk)),N=(quantSums(stock.n(stk))))
plot(flqs)+theme_bw()+xlab("Year")+facet_wrap(~qname,scales="free_y")

```

\pagebreak

### Plot SS3 Stock Dynamics


```{r fig1.1.2, fig.height=8,fig.width=6, fig.cap = paste0("Stock assessment trajectories at age")}

plotdyn(stk)
```

\newpage 

```{r fig1.1.3, fig.height=6, fig.cap = paste0("Stock biology trajectories at age")}

plotbioyr(stk)+
  ggtitle(paste0(stk@name))

```

\newpage 

```{r fig1.1.4, fig.height=6, fig.cap = paste0("Annual stock quanties at age")}

plotbioage(stk)+theme(legend.position = "none")
```



\pagebreak

## Consistency checks using backcasting

Set seed

```{r}
set.seed(1507)
```

Get bias adjusted recruitment deviations from ss3 model


```{r}
stk1 = FLCore::simplify(stk)
out = ss3om::readOutputss3(run)
yrs = an(dimnames(stk)$year)
recruit = out$recruit[out$recruit$Yr%in%yrs,]
dms <- list(year = yrs, season = out$spawnseas)
sigR =  an(out$sigma_R_info[1,3]) # Realised sigR
residuals <- FLQuant(exp(recruit$dev - 0.5 * recruit$biasadjuster *sigR^2), 
        dimnames = c(age = 0, dms), units = "")
residuals[is.na(residuals)] =1# exp(-0.5*sigR^2)


```

```{r}
# TUNE
for(i in 1:4){
testF = fwd(stk1, sr=sr,
  control=fwdControl(year=1984:2023, value=unitMeans(fbar(stk1)[, ac(1984:2023)]),
  quant="fbar"),deviances=residuals)
residuals= residuals*exp(log(rec(stk1))-log(rec(testF)))

testF = fwd(stk1, sr=sr,
  control=fwdControl(year=1984:2023, value=unitMeans(fbar(stk1)[, ac(1984:2023)]),
  quant="fbar"),deviances=residuals)
}
```

```{r hc1, message=FALSE,warning=FALSE,fig.width=9,fig.height=7, fig.cap = paste0("Comparison of stock trajectories from ss3om and backcasting") }

testF = fwd(stk1, sr=sr,
  control=fwdControl(year=1984:2023, value=unitMeans(fbar(stk1)[, ac(1984:2023)]),
  quant="fbar"),deviances=residuals)

testC = fwd(stk1, sr=sr,
  control=fwdControl(year=1984:2023, value=catch(stk1)[, ac(1984:2023)],
  quant="catch"),deviances=residuals)

plot(window(FLStocks(ss3om=stk1,backtestF=testF,backtestC=testC),start=1983))+theme_bw()+facet_wrap(~qname,scale="free_y")

```

