---
title: "F-based forcasting for the annual multi-fleet Stock Synthesis Model of blackspot seabream through 2023"

author: 
 - Henning Winker (GFCM) 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    number_sections: yes
    #toc: yes
    #keep_tex: false
vignette: >
  %\VignetteIndexEntry{ss3daigs Vignetted}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r wrap-hook, echo = FALSE,message=FALSE,warning=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
```


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "  " ,fig.align = 'center', cache=FALSE,tidy.opts=list(width.cutoff=80), tidy=TRUE)

```

# Background

This vignette introduces some new features for the functions `SSdeltaMVLN` and `SSplotEnsemble` and then uses their utilities for illustrating a worked example for F-based forecasts for the seasonal, multi-fleet model of European hake in the Adriatic Sea (GSAs 17-18)

Forecasting with SS3 is based on so apic F values ($F_{apic}$), whereas the choice of the reference $F$-basis and the associated reference points may differ from the $F_{apic}$ scale. For instance, GFCM and ICES, $F_{bar}$ (option 5) is the default option. It is therefore necessary to rescale $F$-basis to $F_{apic}$ for generating forecasts that are consistent with, e.g., $F_{tgt}$ or $F_{cur}$.     


## Install and load packages

First install the latest versions of `ss3diags` (JABBAmodel branch) and `r4ss` from github

```{r, eval=FALSE, message=FALSE}
devtools::install_github("JABBAmodel/ss3diags")
devtools::install_github("r4ss/r4ss")
```

and load some packages

```{r,message=F,warning=F}
library(r4ss)
library(ss3diags)
```

# SSplotEnsemble with SSdeltaMLVN

## SBR reference run for GSAs 1-3 

All models are annual and sex structured with 9 fleets. 
The $B_{MSY}$ proxy is set to 35\%$SSB_0$ with a corresponding $F_{SB35}$


# F-based forecasting 

$F_{apic}$ is used for good reason in forecasts in order to account for multi- fleet selectivity. Comparing the partial impacts selectivity pattern requires setting the instantaneous rate of fishing mortaly $F$ at comparable constant levels. For this purpose, it is important to consider that the definition of selectivity differs across regions (e.g. $Fbar$ or exploitation rate). With regards to temporal compatibility of partial fleet selectivity effects, $F_{bar}$ has the undesirable property that its scale depends on the pre-specified age range across which $F_a$ is averaged. For example, if $F_{bar}$ is set to ages 2-4 to represent the dominant age classes under the current selectivity regime, but the goal is to evaluate the effect of selecting fish only at age-5, a common $F_{bar}$ would result in disproportionately high $F_a$ on ages 5+. This is because $F_{bar}$ is computed for age ranges that are hardly selected for the definition $S_a$ = $F_a$/$max(F_a)$. For this reason, it is more straight forward to use $F_{apical}$ as the standardized quantity $F$ quantify to account for partial impacts of fleet selectivity. 

In the following, step-by-step guidelines are provided to setup an $F_{apic}$, so that it correctly corresponds to the $F_{bar}$ baseline for $F_{tgt}$ across multiple fleets and seasons.

## Step 1: Basic setup

In this a case, a folder with the reference model run is created and the model outputs are loaded with `r4ss::SS_output`

Define name of reference model folder with the fitted ss3 model

```{r}
model = "basecase"
```

Load reference model 

```{r,eval=F}

ss3rep = SS_output(model)

dir.create("rdata",showWarnings = F)
save(ss3rep,file="rdata/ref2023.rdata")

```

```{r,echo=F}
load("rdata/ref2023.rdata")
```

Next a folder `forecast` is created 

```{r}
dir.create("forecast",showWarnings = F)
```

A new helper function `SSnewrun` was added to `ss3diags` to easily create subfolders for the forecast scenarios. First a `Ftgt` reference folder is created for initial cross-checks 

Specify subfolder path

```{r}
ftgtdir = file.path("forecast","Ftgt")
```

Create new F forecast model folder. Note that the data and control file and ss.exe names need to be specified if these diverge from the defaults `data.ss`, `control.ss` and `ss3.exe`

```{r}
dat = "sbr_dat.ss"
ctl = "sbr_ctl.ss"
ss.exe = "ss3.exe"

SSnewrun <- function(model,dat,ctl,ss.exe="SS.exe",newdir,overwrite=TRUE,pars=TRUE){
  dir.create(path = newdir , showWarnings = F)
  model.run = model
  # Copy files
  file.copy(file.path(model.run, "starter.ss_new"), file.path(newdir, "starter.ss"),overwrite=overwrite)
  file.copy(file.path(model.run, ctl), file.path(newdir, ctl),overwrite=overwrite)
  file.copy(file.path(model.run, dat), file.path(newdir, dat),overwrite=overwrite)
  file.copy(file.path(model.run, "forecast.ss"), file.path(newdir, "forecast.ss"),overwrite=overwrite)
  file.copy(file.path(model.run, ss.exe), file.path(newdir, ss.exe),overwrite=overwrite)
  # Automatically ignored for models without wtatage.ss
  file.copy(file.path(model.run, "wtatage.ss"), file.path(newdir, "wtatage.ss"),overwrite=overwrite)
  # add parameters
  if(pars) file.copy(file.path(model.run, "ss.par"), file.path(newdir, "ss.par"),overwrite=overwrite)
} # end function


SSnewrun(model=model,dat=dat,ctl=ctl,newdir=ftgtdir,ss.exe="ss3.exe")
```

Now the forecast file can be read be read with `r4ss`

```{r}
fc <- SS_readforecast(file.path(ftgtdir, "forecast.ss"),verbose = F)
```

## Step 2: Initial F exploitation calculations for Fapic forecast 

Extract the `$exploitation` output from the report file

```{r}
Fexp = ss3rep$exploitation
```

Importantly, the `annual_F` are scaled to the F-basis (here $F_{bar}$), whereas fleet specific $F$ values are always given as $F_{apic}$

Next compute the combined $F_{apic}$ generically across fleets

```{r}
Fexp$Fapic = apply(as.matrix(ss3rep$exploitation[,-c(1:6)]),1,sum,na.rm=T)
```

and aggregate across seasons, by taking the `mean` and not the `sum`.

```{r}
Fapic = aggregate(Fapic~Yr,Fexp,mean)
```

Next compute the corresponding annual $F_{bar}$ values from the `annual_F` 

```{r}
Fbar = aggregate(annual_F~Yr,Fexp,mean)
```

To work out exact ratio between $F_{apic}$ and $F_{bar}$ so that it is consistent with the benchmark calculations with ss3, it is necessary to extract the reference years for selectivity from the `forecast.ss` file. 

The information required for the average selectivity conditions can be found in the `forecast.ss` file under `$Bmark_years`. The third and fourth position define the time horizon for the average selectivity across fleet, a value of -999 (here) indicates that the whole time series is use, but more commonly averages are taken, e.g. over the last 3 years, which can be specified as -2 0 or 2019 2021. The following code attempts to compute this generically.


```{r}
endyr = ss3rep$endyr
if(fc$Bmark_years[3]< -90){
  nfc = length(min(ss3rep$exploitation$Yr+1):endyr) # excluded init year
} else { # if specified (e.g. -2, 0)
  nfc = fc$Bmark_years[4]-fc$Bmark_years[3]+1 
}
# Benchmark reference years
bmyrs = (endyr-nfc+1):endyr
```

```{r}
Fratio = mean(Fapic$Fapic[Fapic$Yr%in%max(bmyrs)]/Fbar$annual_F[Fbar$Yr%in%max(bmyrs)])
Fratio
```
`Fratio` defines the ratio of $F_{apic}$ to $F_{bar}$ for the reference period 

Get the $F_{tgt}$ reference point, here defined as $F_{B_{35}}$. Therefore, the `annF_Btgt` is extracted.

```{r}
Fref = c("annF_Btgt","annF_MSY","annF_SPR")[1] 
Ftgt = ss3rep$derived_quants$Value[ss3rep$derived_quants$Label==Fref]
Ftgt.apic = Ftgt*Fratio
Ftgt  # Fbar
Ftgt.apic
```
## Setting up the manual F forecast input structure

First, do some basic house keeping for the model structure

```{r}
nseas = length(unique(ss3rep$exploitation$Seas)) # number of seasons
fleets = unique(ss3rep$fatage$Fleet) # fleets
nfleets = length(fleets) # number of fleet
```

Next, the mean Fapic by fleet and season is calculated 

```{r}
# subset to benchmark years for selectivity
fexp = ss3rep$exploitation[ss3rep$exploitation$Yr%in%bmyrs,] 
fexp = cbind(fexp[,1:2],fexp[,-c(1:5)])[,-3] #><>  single fleet trick
# flip
fexp = reshape2::melt(fexp, id.vars = c("Yr", "Seas"),
                      variable.name = "Fleet", 
                      value.name = "Fapic")
head(fexp)
```
The forecast file requires Fleet IDs not names. In the next step these are extracted and fleet names are converted in to Fleet IDs 

```{r}
fleet = data.frame(Fleet =ss3rep$FleetNames,ID=ss3rep$fleet_ID)
fexp$Fleet= fleet[match(fexp$Fleet,fleet$Fleet),2]
```

Then, the relative proportions of $F_{apic}$ by fleet and season can be computed

```{r}
Fap = aggregate(Fapic~Seas+Fleet,fexp,mean)
Fap$prop = Fap$Fapic/sum(Fap$Fapic)*nseas
Fap
```

In the next step, status quo $F_{sq}$ for forecasting over the intermediate year(s) is defined. This can be relatively easily changed to intermediate catch years.  Here, the $F_{sq}$ of the average of the last 3 years is used as average, and the intermediate years are set to 2, account for 1 data lag year and an additional management lag year.


```{r}
# F status q
nfsq = 3 
nint = 1 
```

Compute the $F_{sq}$ as $F_{apic}$ vector by season and fleet 

```{r}

fsq = ss3rep$exploitation[ss3rep$exploitation$Yr%in%(endyr),] 
fsq = cbind(fsq[,1:2],fsq[,-c(1:5)])[,-3]  #><>  single fleet trick
fsq = reshape2::melt(fsq, id.vars = c("Yr", "Seas"),
                     variable.name = "Fleet", 
                     value.name = "Fapic")
Fsq = aggregate(Fapic~Seas+Fleet,fsq,mean)
```

Now, the forecast horizon can be defined in the loaded `starter.ss` object `fc`.

```{r}
fc$Nforecastyrs = 20
nfyrs = fc$Nforecastyrs
fyrs= endyr+c(1:nfyrs) 
```

The F-vector that is passed on the forecast file comprises the season/fleet structure replicates for `ninit` for $F_{sq}$ and the forecast years under  $F_{tgt}$ that is scaled to $F_{apic}$ by the `Fratio` and partioned by season and fleets.

```{r}
fvec = c(rep(Fsq$Fapic,nint),rep(Ftgt*Fratio*Fap$prop,nfyrs-nint))
```

Given the fleet, season, intermediate year and forecast years structures, the forecast table for the `forecast.ss` file can finally be constructed.

```{r}
fc$ForeCatch = data.frame("Year"=rep(fyrs,each=nseas*nfleets),"Seas"=1:nseas,
                          "Fleet"=rep(fleets,each=nseas),
                          "Catch or F"=fvec,
                          "Basis"=99)
head(fc$ForeCatch,9)
```                          
Note that the `Basis` 99 specifies that $Fs$ are inputted (2 would be Catch).
Finally, the forecast options need to be adjusted for manual input

```{r}
fc$eof=TRUE 
fc$InputBasis = -1
```

and then the modified `starter.ss` file can be saved 

```{r}
SS_writeforecast(fc, file =file.path(ftgtdir, "forecast.ss"),overwrite = T)
```

## Running Ftgt forecasts with checks

In principle, the `Ftgt` can serve as a reference and the model does not have to be run if the goal is set up a number forecasts relative to $F_{tgt}$.

However, for illustration, the `Ftgt` forecast is run to check that the $F_{apic}$ will produce $F_{bar}$ estimates that are consistent with $F_{tgt}$.

To run

```{r,eval=F}
r4ss::run(ftgtdir, skipfinished = F, show_in_console = T, exe="ss3.exe")
```

After the run is finished (here under 3 min) the output can be loaded again.

```{r,eval=F}
ftgtrep = SS_output(ftgtdir)
save(ftgtrep,file="rdata/ftgtref.rdata")

```

```{r,echo=F}
load("rdata/ftgtref.rdata")
```

For testing, `SSdeltaMVLN` is used to computed the trajectories with CIs, but this time the option `addprj=T` needs to be added to also include the forecast years

```{r}
ftgt.test =SSdeltaMVLN(ftgtrep,Fref="Btgt", 
                       run="Ftgt",addprj = T,plot=F,verbose=F)


```


```{r p5, fig.height=10,fig.width=8, fig.cap = paste0("Stock trajectories with 95\\% CIs for basecase run and a $F_{tgt}$ forecast ")}

sspar(mfrow=c(3,2),plot.cex = 0.8)
  SSplotEnsemble(ftgt.test, add=T,verbose=F,legendloc = "topleft")
```

It can be readily seen that the $F_{apic}$ based $F_{tgt}$ forecast corresponds indeed to the $F_{tgt}$ estimate on $F_{bar}$ scale.  

```{r,warning=FALSE}
library(FLRef)
mvn = FLRef::ssmvln(ftgtrep,Fref="Btgt",addprj = T,verbose=F)

stkf = ss2FLStockR(mvn)
rps = stkf@refpts
stkf@refpts = rbind(rps,
                    FLPar(Bpa=rps["Btgt"]*0.5,
                          Blim=rps["Btgt"]*0.25))
```



```{r p5a, message=F,warning=F,fig.height=7.5,fig.width=8, fig.cap = paste0("Stock trajectories for basecase run and a $F_{tgt}$ forecast, relative to reference points")}

plotAdvice(stkf)+geom_vline(xintercept = 2022.5,linetype=2)
```

```{r p6, fig.height=4,fig.width=6, fig.cap = paste0("Stock trajectories with 95\\% CIs for basecase run and a $F_{tgt}$ forecast")}

sspar(mfrow=c(1,1),plot.cex = 0.8)
  SSplotEnsemble(ftgt.test, add=T,
                 subplots="harvest",verbose=F,
                 legendloc = "topleft",ylabs=expression(F/F[SB35]))
```

\pagebreak

## Looping through forecast scenarios

Set up ratios relative to $F_{sq}$ in this case

```{r}
Ffrac = c(0.001,0.25,0.5,0.75,1,1.25)
```

Specify forecast folders

```{r}
fcdirs = file.path("forecast",paste0("Fsq",Ffrac))
```

Loop through the process of modifying the `forecast.ss` file iteratively. 
The `Ffrac` is applied to apportioned $F_{tgt}$ vector.

```{r,eval=F}

# TODO add optopm for parallel computing 
for(i in 1:length(Ffrac)){
  # create model folder
  SSnewrun(model=ftgtdir,dat=dat,ctl=ctl,newdir=fcdirs[i],ss.exe = "ss3.exe")
  # Read Forecast file
  fc <- SS_readforecast(file.path(fcdirs[i], "forecast.ss"))
  # Apply Ffrac
  # Create F forecast vector (generic)
  # Change to Fsq
  fvec = c(rep(Fsq$Fapic,nint),rep(Fsq$Fapic*Ffrac[i],nfyrs-nint))
  # Creat F forecast table in forecast.ss
  fc$ForeCatch = data.frame("Year"=rep(fyrs,each=nseas*nfleets),"Seas"=1:nseas,
                            "Fleet"=rep(fleets,each=nseas),
                            "Catch or F"=fvec,
                            "Basis"=99) 
  SS_writeforecast(fc, file =file.path(fcdirs[i], "forecast.ss"),overwrite = T)
  r4ss::run(fcdirs[i], skipfinished = F, show_in_console = TRUE, exe=ss.exe)
}
```

Load all runs in one go with `SSgetoutput`

```{r,eval=F}
fcs = SSgetoutput(dirvec = c(fcdirs,ftgtdir))
save(fcs,file="rdata/forecasts.rdata")
```

```{r,echo=F}
load("rdata/forecasts.rdata")
```

Create list with outputs from `SSdeltaMVLN`


```{r p8, fig.height=10,fig.width=8, fig.cap = paste0("Kobe plot checks for the final forecast years")}

sspar(mfrow=c(3,3),plot.cex = 0.8)

fmvns = Map(function(x,y){
  out = SSdeltaMVLN(x,Fref="Btgt",verbose=F,run=y,addprj=T,plot=T)
  mtext(y,outer = F,cex=0.8)
  return(out)
},x=fcs,y=as.list(c(paste0("Fsq",Ffrac),"Ftgt")))

```

Plot forecasts with `SSplotEnsemble`


```{r p9, fig.height=10,fig.width=8, fig.cap = paste0("Stock trajectories with 95\\% CIs for basecase run and forecast scanarios relative $F_{sq}$ and for $F_{tgt}$ ")}

sspar(mfrow=c(3,2),plot.cex = 0.8)
SSplotEnsemble(fmvns, add=T,verbose=F,legendloc = "topleft")

```

\pagebreak

Convert to `FLR` format


```{r}

fstks = FLStocks(Map(function(x,y){
  out = FLRef::ssmvln(x,Fref="Btgt",verbose=F,run=y,addprj=T)
  out = ss2FLStockR(out)
  out@refpts =  stkf@refpts[-4] # Remove B0
  return(out)
},x=fcs,y=as.list(c(paste0("Fsq",Ffrac),"Ftgt"))))
names(fstks) = c(paste0("Fsq",Ffrac),"Ftgt")

```

\pagebreak

```{r p10, fig.height=6.5,fig.width=9, warning=F,message=F, fig.cap = paste0("Trajectories for the basecase run and forecast scanarios relative $F_{sq}$ and for $F_{tgt}$ ")}

plotAdvice(window(fstks))+geom_vline(xintercept = c(2023,2024),linetype=28)

```

\pagebreak

```{r}
out =cbind(FLRef::fwd2stars(fstks,eval.yrs = 2026),
           FLRef::fwd2stars(fstks,eval.yrs = 2026,rel=T)[,3:4])

out[,2:3] = round(out[,2:3],1) 
out[,4:6] =  round(out[,4:6],3)
knitr::kable(out,"pipe",
      align ="lccccc",caption="Summary of short-term forecast scenario results for 2025")

```


```{r}
out =cbind(FLRef::fwd2stars(fstks,eval.yrs = 2043),
           FLRef::fwd2stars(fstks,eval.yrs = 2043,rel=T)[,3:4])

out[,2:3] = round(out[,2:3],1) 
out[,4:6] =  round(out[,4:6],3)
knitr::kable(out,"pipe",
      align ="lccccc",caption="Summary of long-term forecast scenario results for 2043")

```
